---
title: "Pivoting Google Sheets spreadsheets with R"
output: github_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
googlesheets4::gs4_auth(email = TRUE)
```

## Summarising our data

Often, we must summarise our spreadsheets with the help of [pivot tables](https://support.google.com/docs/answer/1272900?hl=en&co=GENIE.Platform%3DDesktop).  Pivot tables are designed to analyze and summarize data, but they rely on data being structured in a specific way to do so effectively.  Pivot tables require data to be in a **long-form format**, where each observation is a separate row, and each variable is a separate column.  However, data is often entered in **wide-form format** for ease of data entry.

If the data is in wide-form format rather than long-form format, we will encounter issues like this when we create pivot tables:

<div style="text-align: center;">
  <video width="640" height="480" controls>
    <source src="untidy.mp4" type="video/mp4">
  Your browser does not support the video tag.
  </video>
</div>

We're unable to easily break down the yearly average market share for each vendor, because each vendor is spread widely over many columns.  Additionally, our analysis is inflexible because it's extremely time consuming and error-prone to switch to a different summary, like the yearly max market share for each vendor

To proceed to summarise our data, we need our data to be in long-form, where those vendors belong to a single column.

## Wide and long data, and some reason for why they exist

Wide-form data is convenient for data entry, which is what we often do in spreadsheets.  However, when the time comes to summarise the data, we may find ourselves at an impasse.

Long-form data provides a structured and organized format that is ideal for data analysis and manipulation. In addition to being essential for pivot tables, long-form data is easier to work with and less prone to errors than wide-form data. With long-form data, each observation is a separate row, and each variable is a column, making it easier to compute with because many environments expect long form data.

Wide-form is useful for a few things, such as cross tables and data entry, but it's not suitable for pivot tables. By understanding the importance of long-form data and using R's tidyverse to manipulate and transform the data, users can create more effective pivot tables and make better-informed decisions.

The long-form format is essential for creating pivot tables in Google Sheets.

![[Tidy](https://www.jstatsoft.org/article/view/v059i10) long-form data vs wide-form data.](tidy.png)


## Why we'll pivot from wide to long with R rather than Sheets

As the data matures, it's normal to evolve from wide-form data (in the data entry phase), to long-form data (in the data summarisation phase).  So then, it is disconcerting that Google Sheets is woefully unfit for the task of pivoting from wide to long.

Google Sheets offers some basic features for data transformation, but it can be inadequate for transforming data from a wide-form to a long-form format. One of the main reasons is that Google Sheets lacks automated tools for data cleaning, transformation, and manipulation.

Converting data from a wide-form to a long-form format can be a time-consuming and error-prone process when done manually with Sheets. For example, it often involves copying and pasting data, manually creating new columns and rows, and deleting unwanted data. These manual processes can be cumbersome, especially when dealing with large and complex datasets.

In contrast, R's [tidyverse](https://tidyverse.org) provides a suite of powerful tools for data transformation and manipulation. The [tidyr](https://tidyr.tidyverse.org/) package, in particular, offers functions for converting data between wide-form and long-form formats, making it much easier to restructure data for analysis. The package also offers tools for cleaning, filtering, and transforming data, allowing users to automate many of the manual processes that can be time-consuming and error-prone in Google Sheets.


## The googlesheets4 library

The [googlesheets4](https://googlesheets4.tidyverse.org/) library is an R package in the tidyverse that provides a simple and powerful interface for interacting with Google Sheets directly from R. With googlesheets4, users can easily read, write, and manipulate data in Google Sheets without leaving the R environment.

One of the key features of googlesheets4 is its support for the latest Google Sheets API.  With this support, users can easily import data from Google Sheets into R for further analysis, using tools from tidyr and [dplyr](https://dplyr.tidyverse.org/) to manipulate the data as needed.

Once the data has been imported into R, users can use the full range of R's data manipulation tools to clean the data. What we're interested in here is pivoting from wide-form to long-form, which allows users to summarize data by different variables and attributes. With googlesheets4, users can easily pivot data in R and then export the results back to Google Sheets.


## Install and load packages.
To install googlesheets4 and other tidyverse packages, input `install.packages("tidyverse")` at your R session's command prompt.

To load googlesheets4, and the R package's we'll use in this session, input this similarly:

```{r load}
library(googlesheets4)
library(tidyverse)
```


## The workflow


### Import from Sheets to R

Here is the spreadsheet we'll be pivoting from wide to long: [<link>](https://docs.google.com/spreadsheets/d/1QG_9gkWtciEmHggb6BjLg-6hF9FyekzKN9cLjyteh00/edit#gid=1469251738&fvid=2032560885)

With the `read_sheet` function, users can easily specify the name of the Google Sheet they want to import data from and the range of cells they want to import. This makes it easy to work with large and complex datasets in R, without the need to manually download and import the data into R. Additionally, the `read_sheet()` function supports various data types and automatically detects and converts data types as needed.

In your analyses, you'll often need to tweak your inputs a bit.  In this case, we instruct the program to read from the data located in the first worksheet in the workbook, in the table ranging over B3 to Z123.

```{r import}
URL_untidy <- "https://docs.google.com/spreadsheets/d/1QG_9gkWtciEmHggb6BjLg-6hF9FyekzKN9cLjyteh00/edit#gid=1469251738&fvid=2032560885"

df <- read_sheet(URL_untidy, sheet = 1, range = "B3:Z123")

df
```

The imported dataframe is in wide-form format.

### Pivot the data in R to long-form

To pivot the data imported with googlesheets4, users can use the tidyr package, which offers a range of functions for manipulating data in long-form and wide-form formats. For example, users can use the [` pivot_longer()`](https://tidyr.tidyverse.org/reference/pivot_longer.html) function to convert wide-form data to long-form data, making it easier to work with in R. Users can also use the [`pivot_wider()`](https://tidyr.tidyverse.org/reference/pivot_wider.html) function to pivot the data back to wide-form.

```{r pivot}
df <- df |> 
  pivot_longer(
    cols      = 4:ncol(df),
    names_to  = "Vendor",
    values_to = "Market Share"
  )

df
```
With only a few lines of code, the data has automatically been made long.  This data is now suitable for a pivot table.

### Export from R to Sheets

Once the data has been transformed in R, users can use the write_sheet() function in googlesheets4 to export the data back to Google Sheets. This allows them to share the results with other team members or stakeholders and collaborate on further analysis or reporting.

We begin by creating a destination spreadsheet.

```{r export-1}
gs4_create("tidy_spreadsheet")
```

After locating the spreadsheet's address, we record the address and then export to that address.

```{r export-2}
URL_tidy <- "https://docs.google.com/spreadsheets/d/11dzJ6Mff6eFhl5h6QO0NK2PYDs8zmutvSh1qX-PFe7U/edit#gid=0"
write_sheet(data = df, ss = URL_tidy, sheet = 1)
```

### Outcome and summary

The data's completely tidy, and we can now easily summarise it to compare average market share with pivot tables

<div style="text-align: center;">
  <video width="640" height="480" controls>
    <source src="tidy_sheet.mp4" type="video/mp4">
  Your browser does not support the video tag.
  </video>
</div>


Overall, the googlesheets4 library provides an accessible roue for transforming wide-form data into long-form data in Google Sheets. Its support for the latest Google Sheets API and seamless integration with R's powertools make it an ideal choice for the multitude of people who need to work with Google Sheets data.

### References & Links

* Tidy: https://www.jstatsoft.org/article/view/v059i10 - link to an article comparing long-form data vs wide-form data.
* Tidyverse: https://tidyverse.org - link to the tidyverse website, which provides a suite of powerful tools for data transformation and manipulation.
* tidyr: https://tidyr.tidyverse.org/ - link to the tidyr package, which offers functions for converting data between wide-form and long-form formats.
* dplyr: https://dplyr.tidyverse.org/ - link to the dplyr package, which provides tools for cleaning, filtering, and transforming data.
* googlesheets4: https://googlesheets4.tidyverse.org/ - link to the googlesheets4 package, which provides a simple and powerful interface for interacting with Google Sheets directly from R.
* Google Sheet example: https://docs.google.com/spreadsheets/d/1QG_9gkWtciEmHggb6BjLg-6hF9FyekzKN9cLjyteh00/edit#gid=1469251738&fvid=2032560885 - link to a Google Sheet that will be used in the tutorial for pivoting from wide to long in R.
* pivot_longer(): https://tidyr.tidyverse.org/reference/pivot_longer.html - link to the documentation for the pivot_longer() function in the tidyr package.
* pivot_wider(): https://tidyr.tidyverse.org/reference/pivot_wider.html - link to the documentation for the pivot_wider() function in the tidyr package.

<!-- ## Including Code -->

<!-- You can include R code in the document as follows: -->

<!-- ```{r cars} -->
<!-- summary(cars) -->
<!-- ``` -->

<!-- ## Including Plots -->

<!-- You can also embed plots, for example: -->

<!-- ```{r pressure, echo=FALSE} -->
<!-- plot(pressure) -->
<!-- ``` -->

<!-- Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot. -->
